# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r echo = TRUE}
data <- read.csv("activity.csv", header=T)
```

data subset without NA values (may not need it due to the above)

```{r echo = TRUE}
subsetData <- subset(data, !is.na(data$steps)) ; head(subsetData)
```

## What is mean total number of steps taken per day?

get the mean of the data
```{r summary.ststs,echo=TRUE}
summary.stats <- ddply(data, .(date), summarize,
                sum.steps = sum(steps, na.rm = T),
                mean.steps = mean(steps, na.rm = T),
                median.steps = median(steps, na.rm = T)
            )
```
Histogram:
```{r getMeanAndMedian, echo = TRUE}

hist(summary.stats$sum.steps, 
     main="Histogram of number of steps per day",
     xlab="Number of steps in a day",
     ylab="Frequency (number of days)")

meanSteps <- mean(subsetData$steps)
medianSteps <- median(subsetData$steps)

```

The _mean_ number of steps is: **`r meanSteps`**.
The _median_ number of steps is: **`r medianSteps`**.

**Note: the section above needs to be repeated**




## What is the average daily activity pattern?



## Imputing missing values

```{r missingData echo = TRUE}
missingSteps <- length( data$steps[ is.na(data$steps) ] )
```

The total number of missing values is: **`r missingSteps`**.

It is assumed that missing values represent periods of inactivity where 

```{r dataimputation, echo=TRUE} 
    dataImputed <- data
    dataImputed$st <- gsub()
```

## Are there differences in activity patterns between weekdays and weekends?

```{r echo = TRUE}


```