# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

Load the data into R. Making sure that the **header=T** option is specified to make sure that the column headings are imported properly, meaning that the structure of the entire imported dataset is intact.

```{r echo = TRUE}
data <- read.csv("activity.csv", header=T)
```



## What is mean total number of steps taken per day?

<explain how you made the histogram>

Summarise the data by the sum, mean and median of the number of steps measured on each day: 

```{r summaryStatsDaily,echo=TRUE}
library(plyr)
dailySummaryStats <- ddply(data, .(date), summarize,
                sum.steps = sum(steps, na.rm = T),
                mean.steps = mean(steps, na.rm = T),
                median.steps = median(steps, na.rm = T)
            )
```


Histogram of the total number of steps measured per day:
```{r getMeanAndMedian, echo = TRUE}

    hist(dailySummaryStats$sum.steps, 
         main="Histogram: Total number of steps per day",
         xlab="Total number of steps in a day",
         ylab="Frequency (number of days)")

    meanSteps <- mean(dailySummaryStats$sum.steps)
    medianSteps <- median(dailySummaryStats$sum.steps)

```

The _mean_ number of steps is: **`r meanSteps`**.  
The _median_ number of steps is: **`r medianSteps`**.


## What is the average daily activity pattern?

<span style="color:red" >Section NOT finished!!<span>

**questions asked**

Creating the time series plot of the **5-minute interval** (x-axis) and the **average** number of steps taken, averaged across all days (y-axis)

```{r plotAvgDailyActivityPattern,echo=TRUE}
    library(plyr)
    intervalSummaryStats <- ddply(data, .(interval), summarize,
                sum.steps = sum(steps, na.rm = T),
                mean.interval.steps = mean(steps, na.rm = T)
                )

    plot(intervalSummaryStats$interval, intervalSummaryStats$mean.interval.steps, type="l",
         main="Plot of Daily Average Activity pattern",
         xlab="Daily interval",
         ylab="Average number of steps per interval"
         )
```

Calculating the 5-minute interval that contains the maximum number of steps, on average across all the days in the dataset. 

Put another, finding out which of the daily intervals had the highest average number of steps (averaged across the **61 days** represented in the original data set).  

```{r getHighestInterval}
# what is the highest value of the mean daily interval?
    maxIntervalMean <- max(intervalSummaryStats$mean.interval.steps)
    
# what interval does this value correspond to?
    targetRow <- subset(intervalSummaryStats, mean.interval.steps == maxIntervalMean)
    maxInterval <- targetRow$interval
```

The interval with the highest mean is **`r maxInterval`**, which has a mean of **`r maxIntervalMean`**.  

## Imputing missing values

```{r missingData, echo = TRUE}
    missingSteps <- length( data$steps[ is.na(data$steps) ] )
```

The total number of missing values is: **`r missingSteps`**.

***<span style="color:red">I have changed this!!</span>*** 

After some consideration, I felt that the simplest way to impute missing values into the data was to replace the NA values with 0. This is based in the simple assumptin that where the data could not be measured/obtained (i.e. was ***NA***), this was because there was no activity to be measured (i.e. the number of steps take were ***zero***). Such a situation may arise in scenario such as when the subject is in the act, sitting, standing still, lying down or traveling in a vehicle. 

Side note: Alternatively, I could take the mean of the "steps" column in the raw data. Which would average out the number of steps taken in each interval of every day. But this may assume constant motion on the part of the subject at all times, not so feasible.

```{r dataimputation, echo=TRUE} 
    # Create a variable called imputedSteps from the steps column of the "data" data frame 
    imputedSteps <- data$steps
    
    # replace the NA values from imputedSteps with the desired value to use for imputation
    # imputation value = 33
        # representing a low activity day)
        # generated from calling summary() on the "data" object. i.e. summary(data).

    imputedSteps[is.na(imputedSteps)] <- 33

    # make a new data frame called imputedData:
        # from imputedSteps variable and the "date" and "interval" columns of "data"
        # using data.frame() to ensure that R makes a data frame and NOT a matrix
    imputedData <- data.frame(imputedSteps, data$date, data$interval)
    
    # rename the column names of "imputedData" data frame to match those of "data"
    names(imputedData) <- names(data)
    
    # make a summary of the data via the method used earlier. This time we split the data by
    # the 
    library(plyr)
    imputedDailySummaryStats <- ddply(imputedData, .(date), summarize,
                sum.steps = sum(steps, na.rm = T),
                mean.interval.steps = mean(steps, na.rm = T)
            )
    # plot the data
        hist(imputedDailySummaryStats$sum.steps, col="red",
         main="Histogram: Total number of steps per day",
         xlab="Total number of steps in a day",
         ylab="Frequency (number of days)")

    imputedMeanSteps <- mean(imputedDailySummaryStats$sum.steps)
    imputedMedianSteps <- median(imputedDailySummaryStats$sum.steps)
```

Temp note: try running something like format to make the printing tidy

In the imputed data:  
1. The _mean_ number of steps is: **`r imputedMeanSteps`**.  
2. The _median_ number of steps is: **`r imputedMedianSteps`**.  

In the data summary of the date generated before the imputation:  
1. The _mean_ number of steps is: **`r meanSteps`**.  
2. The _median_ number of steps is: **`r medianSteps`**.  

Comparatively:  
1. The imputed mean values are **`r imputedMeanSteps/meanSteps`** higher than the previously calculated mean.  
2. However, the imputed median values are **the same** before and after imputation.  

This make sense because These results are not different because the operations required to make the **dailySummaryStats** object was created by 

## Are there differences in activity patterns between weekdays and weekends?

<span style="color:red" >Section NOT finished!!<span>

For this part the `weekdays()` function may be of some help here. Use
the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

1. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 


```{r echo = TRUE}

# helpful start: http://www.statmethods.net/management/variables.html

# use this strategy to create / add a new numeric column according to 
# whether the day in data is a weekday or not

    day.type <- ifelse( weekdays(as.Date(imputedData$date) ) == "Saturday" | 
                            weekdays(as.Date(imputedData$date) ) == "Saturday",
                            c(2), c(1)
                           )

# bind this factor to the data frame
    inputFactorData <- data.frame(imputedData, day.type)

# Make the "day.type" column a factor
    inputFactorData$day.type <- factor(inputFactorData$day.type)

# set the levels (categories) of the weekday
    levels(inputFactorData$day.type) <- c("weekday", "weekend")

# now we are all set

```

not sure yet?

## <span style="color:red" >Skipping the time series plots only costs me 5%!!<span>